One-arm spirals in locally isothermal, radially structured self-gravitating discs

### 2014-11-01 paper/model.tex
major updates

\section{Governing equations and numerical models}\label{model} 
The system of interest is a self-gravitating, inviscid fluid disc
orbting a central star of mass $M_*$. We will mainly consider two-dimensional
(2D, or razor-thin) discs in favour of numerical
resolution, but have also carried out some three-dimensional (3D) disc
simulations. Hence, for generality we describe the system in
3D. We use both cylindrical $(R,\phi,z)$ and spherical polar
coordinates $(r,\theta,\phi)$ centred on the star. 
The governing fluid equations in 3D are  
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot\left(\rho\bm{v}\right) =
  0,\label{cont_eq}\\
  &\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bv = -\frac{1}{\rho}\nabla
  p -\nabla \Phi_\mathrm{tot}\label{mom_eq},\\ 
  & \nabla^2\Phi_d = 4 \pi G \rho \label{poisson}, 
\end{align}
where $\rho$ is the mass density, $\bm{v}=(v_r,v_\theta,v_\phi)$ the
fluid velocity, $p$ is the pressure and the total potential
$\Phi_\mathrm{tot} = \Phi_* + \Phi_d$ consists of that from the
central star, 
\begin{align}
  \Phi_*(r) = -\frac{GM_*}{r}, 
\end{align}
where $G$ is the gravitational constant,  and the disc potential
$\Phi_d$. We further assume a locally 
isothermal equation of state 
\begin{align}
  p = c_s^2(R)\rho,
\end{align}
where the sound-speed is given by $c_s=H\Omega_k$, $H=hR$ is the
disc scale-height with constant aspect-ratio $h$ chosen below, and
$\Omega_k=\sqrt{GM_*/R^3}$ is the midplane Keplerian frequency.

We have purposefully adopted the simplest model that was found to
develop one-arm spirals in discs with radial structure. The indirect
potential associated with the non-inertial reference frame is
neglected to avoid complications arising from the motion of the
central star. For the adopted disc masses $M_d\lesssim 0.1 M_*$, this
effect is not expected to be significant. The locally isothermal
equation of state allows us to apply standard linear density wave
theory for results interpretation. 
%This allows us to consider
%an isolated disc.

@@

\section{Diagnostics}

\subsection{Evolution of non-axisymmetirc modes}
The disc evolution is quantified using mode amplitudes and angular
momenta as follows. We list the 2D definitions with obvious 3D generalisations. 
A hydrodynamic variable $f$ (e.g. $\Sigma$) is written as 
\begin{align}
  f(R,\phi) &= \sum_{m=-\infty}^{\infty}f_m(R)\exp{\ii m \phi} \notag\\
  &= f_0 + 2 \real\left[\sum_{m=1}^\infty f_m \exp{(\ii
      m\phi)}\right], 
\end{align}
where the $f_m$ may be obtained from Fourier transform in $\phi$.

@@

\subsection{Linear density waves}\label{wkb}
In addition to the above diagnostics for disc evolution, we
will also use linear density wave theory to interpret simulation
results.

\subsubsection{Global angular momentum conservation for linear
  perturbations} 
It can be shown that linear perturbations with $\phi$-dependence in the form
$\exp{\ii m\phi }$ satisfies angular momentum conservation
in the form 
\begin{align} 
  \frac{\p \jlin}{\p t} + \nabla\cdot\bm{F} = \tbg, 
\end{align}
where
\begin{align}
  \jlin \equiv
  -\frac{m\Sigma}{2}\imag\left(\bm{\xi}^*\cdot\frac{\p\bm{\xi}}{\p
      t} + \Omega \hat{\bm{k}}\cdot\bm{\xi}\times\bm{\xi}^* + \ii
    m \Omega |\bm{\xi}|^2\right) 
\end{align}
is the angular momentum density, $\bm{\xi}$ is the Lagragian
displacement, and $\bm{F}$ is the angular momentum flux
consisting of a Reynolds stress and a gravitational stress. Explicit
expressions for $\bm{\xi}$ and $\bm{F}$ can be found in e.g.,
\cite{papaloizou85} and \cite{lin11} respectively. In the above
expressions, $\Sigma$ and $\Omega$ correspond to the unperturbed disc.

The term $\tbg$ is 
\begin{align}
  \tbg \equiv
  -\frac{m}{2}\imag\left(\Sigma_m\xi_R^\prime\frac{dc_s^2}{dR}\right), 
\end{align}
and arises because we have adopted a locally isothermal equation of
state. In a barotropic fluid, such as a strictly isothermal disc,
$\tbg$ vanishes and the angular momentum associated with
the perturbation is conserved, provided boundary fluxes vanish.

@@

\subsubsection{Local results}
\cite{shu91} gives the dispersion relation for tightly-wound density
waves of the form $\exp{[\ii(\omega t - m \phi + kR)]}$ in a razor-thin
isothermal disc as  
\begin{align}\label{dispersion}
  (\omega - m\Omega)^2 = \kappa^2 + k^2c_s^2 - 2\pi G \Sigma |k|, 
\end{align}
where $\kappa^2 = R^{-3}\p_R(R^4\Omega^2)$ is the square of the epicycle frequency, 
and $k$ is a real wavenumber such that $|kR|\gg1$.

Given the pattern speed $\Omega_p$,
Eq. \ref{dispersion} can be solved for $|k|$, 
\begin{align}\label{wavenumber}
  \frac{|k|}{k_T} = \frac{2}{Q^2}\left[1 \pm \sqrt{1 -
      Q^2(1-\nu^2)}\right], 
\end{align}
where 
\begin{align}
  k_T \equiv \frac{\kappa^2}{2\pi G \Sigma}
\end{align}
is a characteristic wavenumber, 
\begin{align}
  Q \equiv \frac{c_s\kappa}{\pi G \Sigma}
\end{align}
is the usual Toomre parameter, and
\begin{align}
  \nu \equiv \frac{(\omega - m\Omega)}{\kappa}
\end{align}
is a dimensionless frequency.  The upper (lower) signs in
Eq. \ref{wavenumber} correspond to short (long) waves.

@@

>>>>

\section{Governing equations and numerical models}\label{model} 
The system of interest is a self-gravitating, inviscid fluid disc
orbting a central star of mass $M_*$. We will mainly consider two-dimensional
(2D, or razor-thin) discs in favour of numerical
resolution, but have also carried out some three-dimensional (3D) disc
simulations. Hence, for generality we describe the system in
3D. We use both cylindrical $(R,\phi,z)$ and spherical polar
coordinates $(r,\theta,\phi)$ centred on the star. 
The governing fluid equations in 3D are  
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot\left(\rho\bm{v}\right) =
  0,\label{cont_eq}\\
  &\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bv = -\frac{1}{\rho}\nabla
  p -\nabla \Phi_\mathrm{tot}\label{mom_eq},\\ 
  & \nabla^2\Phi_d = 4 \pi G \rho \label{poisson}, 
\end{align}
where $\rho$ is the mass density, $\bm{v}=(v_r,v_\theta,v_\phi)$ the
fluid velocity, $p$ is the pressure and the total potential
$\Phi_\mathrm{tot} = \Phi_* + \Phi_d$ consists of that from the
central star, 
\begin{align}
  \Phi_*(r) = -\frac{GM_*}{r}, 
\end{align}
where $G$ is the gravitational constant,  and the disc potential
$\Phi_d$. We assume a locally 
isothermal equation of state 
\begin{align}
  p = c_s^2(R)\rho,
\end{align}
where the sound-speed is given by 
\begin{align}
  c_s(R) = hR_0\Omega_k(R_0)\left(\frac{R}{R_0}\right)^{-q/2}, 
\end{align}
where $h$ is the disc aspect-ratio $R=R_0$, 
$\Omega_k=\sqrt{GM_*/R^3}$ is the midplane Keplerian frequency, and
$q$ is the imposed temperature gradient. A strictly isothermal disc
corresponds to $q=0$, and a disc with constant aspect-ratio
corresponds to $q=1$.

% We have purposefully adopted the simplest model that was found to
% develop one-arm spirals in discs with radial structure. 
The indirect potential associated with the non-inertial reference frame is 
neglected to avoid complications arising from the motion of the
central star. For the adopted disc masses $M_d\lesssim 0.1 M_*$, this
effect is not expected to be significant.

% The locally isothermal
% equation of state allows us to apply standard linear density wave
% theory for results interpretation.

%This allows us to consider
%an isolated disc.

@@

\subsection{Diagnostics}

\subsubsection{Evolution of non-axisymmetirc modes}
The disc evolution is quantified using mode amplitudes and angular
momenta as follows. We list the 2D definitions with obvious 3D generalisations. 
A hydrodynamic variable $f$ (e.g. $\Sigma$) is written as 
\begin{align}
  f(R,\phi,t) &= \sum_{m=-\infty}^{\infty}f_m(R,t)\exp{\ii m \phi} \notag\\
  &= f_0(R) + 2 \real\left[\sum_{m=1}^\infty f_m \exp{(\ii
      m\phi)}\right], 
\end{align}
where the $f_m$ may be obtained from Fourier transform in $\phi$.

@@

\section{Linear density waves}\label{wkb}
In this paper we take a fully numerical approach using non-linear
hydrodynamic simulations. However, it will be useful to employ
existing linear density wave theory to interpret the simulations. 
We list here the main definitions and results from 2D linear theory,
taken from \cite{lin11} and \cite{shu91}.

In a linear analysis, one assumes a steady axisymmetric background state, such as
that defined in \S\ref{model}, which is then perturbed such that
\begin{align}  
  \Sigma \to \Sigma(R) + \delta\Sigma_m(R)\exp{\left[\ii\left(-\sigma t +
        m\phi\right)\right]}, 
\end{align}
and similarly for other variables. Since the background is
axisymmetric, we identify $\delta\Sigma_m(R)\exp{\left(-\ii\sigma
    t\right)} = \Sigma_m$ as defined for diagnostics. 
The linearised mass and momentum equations are
\begin{align}
  &-\ii\sbar \delta\Sigma_m = -\frac{1}{R}\frac{d}{dR}\left(R\Sigma\delta
    v_{Rm}\right) - \frac{\ii m \Sigma}{R}\delta v_{\phi m}, \\
  &-\ii\sbar\delta v_{Rm} - 2\Omega \delta v_{\phi m} = -
  c_s^2(R)\frac{d}{dR}\left(\frac{\delta\Sigma_m}{\Sigma}\right) - \frac{d}{dR}\delta\Phi_m,\\
  & -\ii\sbar\delta v_{\phi m} + \frac{\kappa^2}{2\Omega}\delta v_{Rm} =
  -\frac{\ii m }{R}\left(c_s^2\frac{\delta\Sigma_m}{\Sigma} + \delta\Phi_m\right),
\end{align}
where $\sbar = \sigma - m\Omega$ and $\kappa^2 =
R^{-3}\p_R(R^4\Omega^2)$ is the square of the epicycle frequency. A
locally isothermal equation of state has been assumed.    
The linearised Poisson equation is 
\begin{align}
  \nabla^2\delta\Phi_m = 4\pi G \delta\Sigma_m \delta(z). 
\end{align}

The linearised equations can be combined to a single
integro-differential eigenvalue problem. We list below some general
properties of the linear perturbations that will be useful for
understanding our simulations. We defer a full numerical
exploration of the linear problem to a future study.

\subsection{Global angular momentum conservation for linear
  perturbations} 
It can be shown that linear perturbations with $\phi$-dependence in the form
$\exp{(\ii m\phi)}$ satisfies angular momentum conservation
in the form 
\begin{align} 
  \frac{\p \jlin}{\p t} + \nabla\cdot\bm{F} = \tbg, 
\end{align}
where
\begin{align}\label{lin_ang_mom_cons}
  \jlin \equiv
  -\frac{m\Sigma}{2}\imag\left(\bm{\xi}^*\cdot\frac{\p\bm{\xi}}{\p
      t} + \Omega \hat{\bm{k}}\cdot\bm{\xi}\times\bm{\xi}^* + \ii
    m \Omega |\bm{\xi}|^2\right) 
\end{align}
is the angular momentum density, $\bm{\xi}$ is the Lagragian
displacement, and $\bm{F}$ is the angular momentum flux
consisting of a Reynolds stress and a gravitational stress. Explicit
expressions for $\bm{\xi}$ and $\bm{F}$ can be found in e.g.,
\cite{papaloizou85} and \cite{lin11} respectively.

The term $\tbg$ is 
\begin{align}\label{baroclinic_torque}
  \tbg \equiv
  -\frac{m}{2}\imag\left(\Sigma_m\xi_R^*\frac{dc_s^2}{dR}\right), 
\end{align}
and arises because we have adopted a locally isothermal equation of
state. In a barotropic fluid, such as a strictly isothermal disc,
$\tbg$ vanishes and the angular momentum associated with
the perturbation is conserved, provided boundary fluxes vanish.

@@

\subsection{Local results}
%The linearised equations correspond to an integro-differential
%equation eigenvalue problem. 
In the local approximation, perturbations are assumed to vary rapidly
relative to any background gradients. \cite{shu91} gives the
dispersion relation for tightly-wound density 
waves of the form $\exp{[\ii(-\sigma t + m \phi + kR)]}$ in a razor-thin
disc as  
\begin{align}\label{dispersion}
  (\sigma - m\Omega)^2 = \kappa^2 + k^2c_s^2 - 2\pi G \Sigma |k|, 
\end{align}
and $k$ is a real wavenumber such that $|kR|\gg1$. Note that in this
approximation, only axisymmetric perturbations with $m=0$ can be
unstable (having an imaginary $\sigma$).

Given the pattern speed $\Omega_p$, Eq. \ref{dispersion} can be solved
for $|k|$,   
\begin{align}\label{wavenumber}
  |k| = \frac{\pi G \Sigma}{c_s^2}\left[1 \pm \sqrt{1 -
      Q^2(1-\nu^2)}\right], 
\end{align}
where 
% \begin{align}
%   k_T \equiv \frac{\kappa^2}{2\pi G \Sigma}
% \end{align}
% is a characteristic wavenumber, 
\begin{align}
  Q \equiv \frac{c_s\kappa}{\pi G \Sigma}
\end{align}
is the usual Toomre parameter, and
\begin{align}
  \nu \equiv \frac{(\omega - m\Omega)}{\kappa}
\end{align}
is a dimensionless frequency. The upper (lower) signs in
Eq. \ref{wavenumber} correspond to short (long) waves.

@@

A detailed discussion of the properties of density waves in the local
approximation can be found in \cite{shu91}. A important result is that
waves interior to co-rotation ($R<R_c$) have negative angular momentum, while
waves outside co-rotation ($R>R_c$) have positive angular
momentum. The local approximation becomes less good for gobal modes,
but will nevertheless be useful in results interpretation.

### 2014-11-01 paper/paper.tex
update
major change in paper - new idea - ang mom exchange with background because of temp gradient
major updates

\newcommand{\p}{\partial}
\newcommand{\mnras}{MNRAS}
\newcommand{\apj}{ApJ}
\newcommand{\aap}{A\&A}
\newcommand{\apjl}{ApJL}
\newcommand{\araa}{ARAA}
\newcommand{\dd}{\delta}
\newcommand{\adot}{\dot{a}}
\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}
\newcommand{\gtrsim}{\;\raisebox{-.8ex}{$\buildrel{\textstyle>}\over\sim$}\;}
\newcommand{\lesssim}{\; \raisebox{-.8ex}{$\buildrel{\textstyle<}\over\sim$}\;}
\newcommand{\anrev}{{\it ARA\&A, }}
\newcommand{\apjs}{{\it ApJS, }}
\newcommand{\icar}{{\it Icarus, }}
\newcommand{\mnr}{{\it MNRAS, }}
\newcommand{\nat}{{\it Nature, }}
\newcommand{\sci}{{\it Sci, }}
\newcommand{\ana}{{\it A\&A, }}
\newcommand{\anas}{{\it A\&AS, }}
\newcommand{\aaps}{{\it A\&AS, }}
\newcommand{\anar}{{\it A\&AR, }}
\newcommand{\prd}{{\it Phys. Rev. D }}
\newcommand{\qjl}{{\it QJRAS, }}
\newcommand{\sbar}{\bar{\sigma}}
\newcommand{\avg}[1]{\langle #1 \rangle_\phi}
\newcommand{\bu}{\bm{u}}
\newcommand{\lmax}{l_\mathrm{max}}
\newcommand{\mmax}{m_\mathrm{max}}
\newcommand{\trel}{t_\mathrm{rel}}
\newcommand{\zeus}{{\tt ZEUS-MP }}
\newcommand{\ii}{\mathrm{i}}
\newcommand{\bv}{\bm{v}}
\newcommand{\rin}{r_\mathrm{in}}
\newcommand{\rout}{r_\mathrm{out}}
\newcommand{\ciso}{c_\mathrm{iso}}
\newcommand{\tbeta}{\tilde{\beta}}
\newcommand{\teta}{\tilde{\eta}}
\newcommand{\tcool}{t_\mathrm{cool}}
\DeclareMathOperator{\erf}{erf}
\DeclareMathOperator{\real}{Re}
\DeclareMathOperator{\imag}{Im}

@@

\title[One-arm spirals in self-gravitating discs]{One-arm spirals in
  self-gravitating dead zones in protoplanetary discs}

@@

\maketitle
\begin{abstract}
  Theoretical models of circumstellar or protoplanetary discs indicate
  the possible existence of `dead zones' --- regions in the disc
  where mass accretion is greatly reduced because of inefficient
  magneto-hydrodynamic turbulence that would otherwise enable 
  accretion. It has been suggested that mass built-up in a dead zone
  may lead to gravitational instability. However, the type of
  gravitational instability relevant in this context has not been
  clarified. In this work, we use numerical simulations to explore
  gravitational instability in astrophysical discs in which only a 
  sub-portion of it is self-gravitating, taken as a toy model for
  protoplanetary disc containing a dead zone. We conduct numerical hydrodynamic
  experiments in 2D using the FARGO code and in 3D with the ZEUS-MP
  and PLUTO codes. In all cases we observe the development of a spiral
  mode with azimuthal wavenumber $m=1$ inside the dead
  zone. Surprisingly, the co-rotation radius of the spiral is well
  outside the dead zone, resulting in a nearly stationary one-arm spiral
  pattern in the dead zone. Our simulations demonstrate the potential
  importance global gravitational interaction between a
  self-gravitating dead zone and the exterior, non-self-gravitating
  parts of the disc.  
\end{abstract}

@@

\input{model}
\input{results}
%\input{discussion}
%\input{summary}

>>>>

\newcommand{\p}{\partial}
\newcommand{\mnras}{MNRAS}
\newcommand{\apj}{ApJ}
\newcommand{\aap}{A\&A}
\newcommand{\apjl}{ApJL}
\newcommand{\araa}{ARAA}
\newcommand{\dd}{\delta}
\newcommand{\adot}{\dot{a}}
\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}
\newcommand{\gtrsim}{\;\raisebox{-.8ex}{$\buildrel{\textstyle>}\over\sim$}\;}
\newcommand{\lesssim}{\; \raisebox{-.8ex}{$\buildrel{\textstyle<}\over\sim$}\;}
\newcommand{\anrev}{{\it ARA\&A, }}
\newcommand{\apjs}{{\it ApJS, }}
\newcommand{\icar}{{\it Icarus, }}
\newcommand{\mnr}{{\it MNRAS, }}
\newcommand{\nat}{{\it Nature, }}
\newcommand{\sci}{{\it Sci, }}
\newcommand{\ana}{{\it A\&A, }}
\newcommand{\anas}{{\it A\&AS, }}
\newcommand{\aaps}{{\it A\&AS, }}
\newcommand{\anar}{{\it A\&AR, }}
\newcommand{\prd}{{\it Phys. Rev. D }}
\newcommand{\qjl}{{\it QJRAS, }}
\newcommand{\sbar}{\bar{\sigma}}
\newcommand{\avg}[1]{\langle #1 \rangle_\phi}
\newcommand{\bu}{\bm{u}}
\newcommand{\lmax}{l_\mathrm{max}}
\newcommand{\mmax}{m_\mathrm{max}}
\newcommand{\trel}{t_\mathrm{rel}}
\newcommand{\zeus}{{\tt ZEUS-MP }}
\newcommand{\ii}{\mathrm{i}}
\newcommand{\bv}{\bm{v}}
\newcommand{\rin}{r_\mathrm{in}}
\newcommand{\rout}{r_\mathrm{out}}
\newcommand{\ciso}{c_\mathrm{iso}}
\newcommand{\tbeta}{\tilde{\beta}}
\newcommand{\teta}{\tilde{\eta}}
\newcommand{\tcool}{t_\mathrm{cool}}
\newcommand{\jlin}{j_\mathrm{lin}}
\newcommand{\jlintot}{J_\mathrm{lin}}
\newcommand{\tbg}{T_\mathrm{BG}}
\DeclareMathOperator{\erf}{erf}
\DeclareMathOperator{\real}{Re}
\DeclareMathOperator{\imag}{Im}

@@

\title[One-arm spirals]{One-arm spirals in 
  locally isothermal, radially structured self-gravitating discs}

@@

\maketitle
\begin{abstract} 
  We present numerical simulations of gravitational instability in
  astrophysical discs with an an imposed temperature profile, and in
  which only a sub-portion of it is self-gravitating. We conduct
  numerical hydrodynamic experiments in 2D using the FARGO code and in
  3D with the ZEUS-MP and PLUTO codes. In such systems we observe the
  development of a 
  spiral mode with azimuthal wavenumber $m=1$ with co-rotation radius well
  outside the bulk of the spiral arm, resulting in a nearly stationary
  one-arm spiral pattern in the self-gravitating region. This one-arm
  spiral does not require the movement of 
  the central star, as found in previous studies. Instead, we find it
  to be associated with a structured disc, as well as an imposed
  temperature gradient. We apply results
  from existing linear theory and attribute the growth of this spiral
  to the non-conservation of angular momentum for linear perturbations
  in discs with an imposed temperature gradient. This enables angular
  momentum exchange between the spiral and the background disc.

  % or confirm in 3D
  %, taken as a toy model of a 
  %protoplanetary disc containing a finite-size massive dead zone.
   % Theoretical models of circumstellar or protoplanetary discs indicate
  % the possible existence of `dead zones' --- regions in the disc
  % where mass accretion is greatly reduced because of inefficient
  % magneto-hydrodynamic turbulence that would otherwise enable 
  % accretion. It has been suggested that mass built-up in a dead zone
  % may lead to gravitational instability. However, what type of
  % gravitational instability is relevant in this context has not been
  % clarified.

  %
  % Our simulations demonstrate the 
  % importance of global gravitational interaction between a 
  % self-gravitating dead zone and the exterior, non-self-gravitating
  % parts of the disc. This non-local effect, due to disc structure, 
  % exists even when the total disc mass is below the threshold for
  % global effects found in
  % previous studies. 
\end{abstract}

@@

\input{model}
\input{results}
%\input{results_3d}
%\input{discussion}
%\input{summary}

### 2014-11-01 paper/results.tex
major updates

\subsection{Reference run with random perturbations}
In this fiducial simulation we subject the disc to initial perturbations in
cylindrical radial velocity, 
\begin{align}\label{randpert}
  \frac{\delta v_R}{c_s} = \frac{\delta}{M}\times T(R) \sum_{m=1}^M\cos{m\phi},
\end{align}
where $\delta$ is an amplitude and 
\begin{align}
  T(R) =
  \exp{\left[-\frac{1}{2}\left(\frac{R-\overline{R}_d}{\Delta
          R_d}\right)^2\right]}, 
\end{align}
where $\overline{R}_d = (R_{d1}+R_{d2})/2$ and $\Delta R_d =
(R_{d2}-R_{d1})/2$. We use $M=10$ and choose
$\delta\in[-10^{-3},10^{-3}]$ randomly but independent of $\phi$.

Fig. \ref{fargo_modeamp} plots evolution of the maximum
non-axisymmetric surface density amplitudes in $R\in[R_{d1},R_{d2}]$
for $m\in[1,10]$. Snapshots from the simulation are shown in
Fig. \ref{fargo_2d}. 
At early times $t\lesssim100P_0$ the dead zone is
dominated by low-amplitude high-$m$ perturbations. The $m\geq4$ modes
growth initially and saturate (or decays) after $t=40P_0$. Notice the
low $m\leq 2$ modes decay initially, but grows between $t\in[20,40]P_0$,
possibily due to saturation of the high-$m$ modes  
\citep{laughlin96,laughlin97}. However, the $m=1$ mode begins to grow
again after $t=70P_0$, and eventually dominates the dead zone.

@@

>>>>

\subsection{Reference run with random perturbations}
In this fiducial simulation we subject the disc to initial perturbations in
cylindrical radial velocity, 
\begin{align}\label{randpert}
  v_R \to v_R+ c_s\frac{\delta}{M}
  \exp{\left[-\frac{1}{2}\left(\frac{R-\overline{R}_d}{\Delta 
          R_d}\right)^2\right]}\sum_{m=1}^M\cos{m\phi},
\end{align}
where $\delta$ is an amplitude, $\overline{R}_d = (R_{d1}+R_{d2})/2$
and $\Delta R_d = (R_{d2}-R_{d1})/2$. We use $M=10$ and choose 
$\delta\in[-10^{-3},10^{-3}]$ randomly but independent of $\phi$.

Fig. \ref{fargo_modeamp} plots evolution of the maximum
non-axisymmetric surface density amplitudes in $R\in[R_{d1},R_{d2}]$
for $m\in[1,10]$. Snapshots from the simulation are shown in
Fig. \ref{fargo_2d}. 
At early times $t\lesssim100P_0$ the disc 
dominated by low-amplitude high-$m$ perturbations. The $m\geq4$ modes
growth initially and saturate (or decays) after $t=40P_0$. Notice the
low $m\leq 2$ modes decay initially, but grows between $t\in[20,40]P_0$,
possibily due to saturation of the high-$m$ modes  
\citep{laughlin96,laughlin97}. However, the $m=1$ mode begins to grow
again after $t=70P_0$, and eventually dominates the annulus.

@@

\subsection{Trapping of the low-frequency $m=1$ spiral}
In order to focus on the $m=1$ spiral that eventually emerges, here 
we examine a simulation initialised with only $m=1$
perturbations, by setting $M=1$ in Eq. \ref{randpert}. The disc
evolution is largely similar to the previous case, but the final spiral is
more co-herent, facilitating the analysis below. 
Fig. \ref{2d_fargo_viz} shows the $m=1$ surface density
distribution at the end of this simulation.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/polarxy2_dens120_fargo.ps}
  \caption{Cartesian visualisation of the $m=1$ surface density
    structure in the FARGO simulation initialised with only $m=1$
    perturbations. 
    \label{2d_fargo_viz}} 
\end{figure}

By examining the disc evolution in $t\in[80,120]P_0$, we find the
one-arm spiral in Fig. \ref{2d_fargo_viz} has a 
co-rotation radius and growth rate of
\begin{align*}
  &R_c \simeq 4.4R_0,\\
  &\gamma\simeq 0.014\Omega_k(R_0) = 0.13\Omega_k(R_c). 
\end{align*}
The spiral can be considered as low frequency because $|\omega|
\lesssim 0.3\Omega$ (recall $\omega = 
m\Omega_p$) in the region it is promininent $R\in[R_{d1},R_{d2}]$. The
growth rate is also slow relative to the local rotation. This gives a
nearly stationary spiral pattern.

%Since most of
%the perturbation lies at $R<R_c$, this mode has negative
%angular momentum according to local linear theory.

Next, we write $\Sigma_1 = 
|\Sigma_1|\exp{(\ii kR)}$, where $k$ is real, and assume the amplitude
$|\Sigma_1|$ varies slowly compared to the complex phase. This is the
main assumption in local theory. We plot in
Fig. \ref{fargo_wavenumber} the normalise wavenumber associated with
the $m=1$ spiral. We find that
\begin{align*}
  kR \sim \frac{\pi G \Sigma}{c_s^2}R \sim \frac{1}{hQ}, 
\end{align*}
where we used $Q\sim c_s\Omega/\pi G \Sigma$ and $R\Omega/c_s\sim
h^{-1}$. Since $Q=O(1)$ and $h\ll 1$, $|kR|\gg 1$ and we 
can apply results from local theory self-consistently. Since $k>0$,
the spiral is trailing, as seen in Fig. \ref{2d_fargo_viz}.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_kr120_fargo.ps}
  \caption{Normalised radial wavenumber of the $m=1$ spiral in
    Fig. \ref{2d_fargo_viz}.\label{fargo_wavenumber}} 
\end{figure}

Using the estimated value of $R_c$, we plot in
Fig. \ref{fargo_qbarrier} the quantity $\nu^2 - 1 + Q^{-2}$, which is
required to be positive in local theory for purely wave-like
solutions to the dispersion relation (Eq. \ref{dispersion}) when the
mode frequency is given. The figure shows two $Q$-barriers located in
the inner disc, at $R_{Qb}=R_0$ and $R_{Qb}=1.6R_0$, which is indeed the
region where the $m=1$ spiral develops. This means that the one-arm
spiral is trapped. There is one outer Lindblad resonance at $R_L\simeq
7.2R_0$. Thus, waves may be launched in $R\gtrsim 7.2R_0$ by the main 
spiral disturbance in the inner disc. %this would be a de-stabilising
% effect.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_Qbar_fargo.ps} 
  \caption{Dimensionless mode frequency $\nu$ for the $m=1$ spiral in
    Fig. \ref{2d_fargo_viz}. The disperion relation for local density
    waves, Eq. \ref{dispersion}, permits purely wave-like solutions in 
    regions where $\nu^2 - 1 + Q^{-2}>0$. 
    \label{fargo_qbarrier}} 
\end{figure}

\subsection{Angular momentum exchange with the background disc} 
The wavenumber $k = \pi G\Sigma/c_s^2$ corresponds to the
wavelength most susceptible to gravitational instability for
\emph{axisymmetric} perturbations. Physically, then, it is not
surprising to find perturbations of this wavenumber in a
self-gravitating disc where $Q\sim 1$. However, according the
local dispersion relation, $m=1$ perturbations are formally stable for 
real $k$.

In order to track down the origin for the (slow) growth of the
$m=1$ spiral, we invoke angular momentum conservation for linear
perturbations, Eq. \ref{lin_ang_mom_cons}. Assuming fluxes are
negligible at the disc boundaries, we can integrate
Eq. \ref{lin_ang_mom_cons} to give
\begin{align}\label{baroclinic_torque_int}
  \frac{d}{dt}\underbrace{\int_{R_\mathrm{min}}^{R_\mathrm{max}}\jlin
    2\pi R dR}_{\jlintot} 
  =\int_{R\mathrm{min}}^{R\mathrm{max}}T_\mathrm{BG} 2\pi R dR, 
\end{align}
where we recall $T_{BG}$ is the torque density related to the imposed
sound-speed profile (Eq. \ref{baroclinic_torque}). We have explicitly
computed both sides of Eq. \ref{baroclinic_torque_int} using
simulation data, and compare them in Fig. \ref{fargo_angmom_ex}.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_ang_fargo.ps} 
  \caption{
    \label{fargo_angmom_ex}} 
\end{figure}

%ang mom decreases slightly more rapidly than torque exchange provides
%- probably non-linear effects/shocks
